---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 2 - Due date 02/05/21"
author: "Abi Vanover"
output: pdf_document
geometry: margin=2.54cm
---

```{r Packges}

pacman::p_load(forecast, tseries, dplyr, lubridate)

```


```{r data work}
#Importing data set

library(readxl)
Df1 <- read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 9)
View(Df1)


```


## Question 1

```{r Q1}

dates <- Df1[2:575, 1]
variables <- Df1[2:575, 4:6]

workingdf <- cbind(dates,variables)

head(workingdf)

biomass <- as.numeric(variables$`Total Biomass Energy Production`)
renewable <- as.numeric(variables$`Total Renewable Energy Production`)
hydro <- as.numeric(variables$`Hydroelectric Power Consumption`)

```

## Question 2


```{r Q2}

#creating each variable as an individual ts to use for plotting later
tsbiomass <- ts(workingdf$`Total Biomass Energy Production`, start = 1973, end = 2020, frequency = 12)

tsrenew <- ts(variables$`Total Renewable Energy Production`, start = 1973, end = 2020, frequency = 12)

tshydro <- ts(variables$`Hydroelectric Power Consumption`, start = 1973, end = 2020, frequency = 12)


#transforming my dataframe into one time series object
tsdf <- ts(workingdf[2:575, 2:4], start = 1973, end = 2020, frequency= 12)

```

## Question 3
Compute mean and standard deviation for these three series.

```{r Q3}
#Trying again for mean & Sd, converting to numeric

x <- mean(as.numeric(tsbiomass))
sd(as.numeric(tsbiomass))

y <- mean(as.numeric(tsrenew))
sd(as.numeric(tsrenew))

z <- mean(as.numeric(tshydro))
sd(as.numeric(tshydro))


```

## Question 4

```{r Q4}
#This is working now 

plot(tsbiomass, xlab = "Time", ylab = "Total Biomass Energy Production")
abline(h=x, col = "dodgerblue")

#This plot is fine and just needs a line
plot(tsrenew, xlab = "Time", ylab = "Total Biomass Energy Production")
abline(h=y, col = "dodgerblue")


#This plot is fine and just needs a line to get added
plot(tshydro, xlab = "Time", ylab = "Total Renewable Energy Production")
abline(h=z, col = "darkcyan")


```

## Question 5

Total biomass and total renewables are significantly positively correlated, with a value of ~.923. This makes sense as total biomass would be a component of total renewables. Renewables and Hydro consumption are not correlated; the value is essentially zero. Biomass production and Hydroelectric power consumption are slightly negatively correlated, which is reasonable as well: as biomass energy production increases, it can be inferred that it would replace some hydroelectric power. 

```{r Q5, echo = FALSE}
#Correlation Tests

cor.test(biomass, renewable)
#.923, Biofuels & Total Biomass

cor.test(renewable, hydro)
#-.00275 

cor.test(biomass, hydro)
#-.256, Biomass & Total Renewables

```

## Question 6

The plots for Biomass and Total Renewables are very similar, with high initial values of autocorrelation that decrease over time. However, the plot for Hydroelectric Power Consumption is significantly different. The autocorrelation plot for Hydro shows several peaks and valleys, with maximums decreasing over time. Several of these values are also negative, which were not present on either of the two previous plots. 


```{r Q6}
acfb=acf(biomass, lag.max = 40, type = "correlation") #decreasing at a shallow slope
acfr=acf(renewable, lag.max = 40, type = "correlation") #smoother decline
acfh=acf(hydro, lag.max = 40, type = "correlation") #decreasing sharper, wave pattern

acfb
acfr
acfh

```

## Question 7

These plots are all very similar to each other, but differ sharply from the initial ACF plots for the Biomass & Total Renewables. For Biomass & Renewables, the ACF plots had a steadily decreasing trendline of all positive values. However, the PACF shows a more cyclical trend of values, with both negative and positive values. For all plots though, the absolute values of the PACF's are lower than the absolute values of ACF's. 


```{r Q7}
pacfb=pacf(biomass, lag.max = 40) #shrinking waves, decreasing over time
pacfr=pacf(renewable, lag.max = 40) #similar pattern, but sharper
pacfh=pacf(hydro, lag.max = 40) #similar pattern, but slightly wider range


pacfb
pacfr
pacfh

```